---
title: "p8105_hw2_yw4886"
author: "Yijia"
date: "2026-10-01"
output: github_document
---
##Set up packages
```{r}
library(tidyverse)
library(readxl)
```

#Problem 1
##Import and clean
```{r}
transit_df =
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```
###Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far
The NYC Transit dataset contains information about subway station entrances and exits, including station line and name, geographic coordinates, routes served, entry status, vending availability, entrance type, and ADA accessibility. I cleaned the data by standardizing the variable names, selecting only the variables relevant to the analysis, and converting the `entry` variable from character values (`YES` and `NO`) to a logical variable (`TRUE` and `FALSE`).

###Give the dimension (rows x columns) of the resulting dataset
```{r}
dim(transit_df)
```
The dataset has 1,868 observations and 19 variables.

###Are these data tidy?
At this stage, the data are not fully tidy because route information is spread across multiple columns (route1 through route11) rather than stored as values of a single route variable.

##Answer the following questions using these data
###How many distinct stations are there?
```{r}
transit_df |>
  distinct(line, station_name) |>
  nrow()
```
###How many stations are ADA compliant?
```{r}
transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
###What proportion of station entrances / exits without vending allow entrance?
```{r}
transit_df |>
  filter(vending == "NO") |>
  summarise(
    proportion = mean(entry)
  )
```
##Reformat data
```{r}
transit_tidy =
  transit_df |>
  mutate(
    across(route1:route11, as.character))|>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name) |>
  mutate(
    route_number = as.integer(route_number)
  )
transit_tidy
```
###How many distinct stations serve the A train? 
```{r}
transit_tidy |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```
###Of the stations that serve the A train, how many are ADA compliant?
```{r}
transit_tidy |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(line, station_name) |>
  nrow()
```
#Problem 2
###Read and clean the Mr. Trash Wheel sheet
```{r}
mr_trash =
  read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N709"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )
mr_trash
```

###Professor Trash Wheel
```{r}
professor_trash =
  read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M134"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Professor Trash Wheel"
  )
professor_trash
```
##Gwynnda
```{r}
gwynnda_trash =
  read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L351"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Gwynnda"
  )
gwynnda_trash
```

##Combine
```{r}
trash_wheel_df =
  bind_rows(
    mr_trash,
    professor_trash,
    gwynnda_trash
  )
trash_wheel_df
```

The combined Trash Wheel dataset contains `r nrow(trash_wheel_df)` observations from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Key variables include dumpster number, month and year of collection, date, weight and volume of trash collected, and counts of different types of trash such as plastic bottles, polystyrene, cigarette butts, plastic bags, and wrappers. During data cleaning, I imported each worksheet using specified Excel ranges to exclude non-data rows and columns, standardized the variable names, removed rows without dumpster-specific information, cleaned the month and year variables, and rounded the number of sports balls in the Mr. Trash Wheel dataset to the nearest integer. I also created a `trash_wheel` variable to identify which Trash Wheel each observation came from before combining the three datasets. Professor Trash Wheel collected a total of `r professor_weight` tons of trash, and Gwynnda collected `r gwynnda_cigarettes` cigarette butts in June 2022.

###For available data, what was the total weight of trash collected by Professor Trash Wheel?
```{r}
professor_weight =
  trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  ) |>
  pull(total_weight)
```

###What was the total number of cigarette butts collected by Gwynnda in June of 2022?
```{r}
gwynnda_cigarettes =
  trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_cigarettes = sum(cigarette_butts, na.rm = TRUE)
  ) |>
  pull(total_cigarettes)
```











