---
title: "p8105_hw2_yw4886"
author: "Yijia"
date: "2026-10-01"
output: github_document
---
##Set up packages
```{r}
library(tidyverse)
library(readxl)
```

#Problem 1
##Import and clean
```{r}
transit_df =
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```
###Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far
The NYC Transit dataset contains information about subway station entrances and exits, including station line and name, geographic coordinates, routes served, entry status, vending availability, entrance type, and ADA accessibility. I cleaned the data by standardizing the variable names, selecting only the variables relevant to the analysis, and converting the `entry` variable from character values (`YES` and `NO`) to a logical variable (`TRUE` and `FALSE`).

###Give the dimension (rows x columns) of the resulting dataset
```{r}
dim(transit_df)
```
The dataset has 1,868 observations and 19 variables.

###Are these data tidy?
At this stage, the data are not fully tidy because route information is spread across multiple columns (route1 through route11) rather than stored as values of a single route variable.

##Answer the following questions using these data
###How many distinct stations are there?
```{r}
transit_df |>
  distinct(line, station_name) |>
  nrow()
```
###How many stations are ADA compliant?
```{r}
transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
###What proportion of station entrances / exits without vending allow entrance?
```{r}
transit_df |>
  filter(vending == "NO") |>
  summarise(
    proportion = mean(entry)
  )
```
##Reformat data
```{r}
transit_tidy =
  transit_df |>
  mutate(
    across(route1:route11, as.character))|>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name) |>
  mutate(
    route_number = as.integer(route_number)
  )
transit_tidy
```
###How many distinct stations serve the A train? 
```{r}
transit_tidy |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```
###Of the stations that serve the A train, how many are ADA compliant?
```{r}
transit_tidy |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(line, station_name) |>
  nrow()
```
#Problem 2









